---
title: "R Course Notes"
author: "Tim Farkas"
date: "5/11/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### The R Console

The R *Console* is an interactive workspace where you can execute R expressions line-by-line in real time.

`>` indicates that a new expression can be executed.

`+` indicates that an expression has not yet been completed.

* Either finish the expression, or press  <kbd>ESC</kbd> to exit.

Each line of executed code is saved to the **history**, and can be accessed quickly using the <kbd>UP</kbd> arrow, or in the History window.

The Console can be cleared with  <kbd>CTL</kbd> + <kbd>L</kbd>.

...

### Objects & Functions

> "Everything that exists in R is an object.  
> Everything that happens in R is a function call"  
> - John Chambers

### Data Types

**Logical** data are `TRUE` or `FALSE`, `T` or `F` -- spelling out whole word recommended.

**Double** data are real-number **numeric** with high maximum and larger memory allocation; default numeric data type.

**Integer** data are whole-number **numeric** with a low maximum and smaller memory allocation; specified with an "L" following a number: `384L`.

**Character**: **strings** of text, including letters, numbers, symbols.  

* A single `character` datum can be a string of many "characters", enclosed in single or double quotes; eg. `"Hello, world!"`.

* double quotes are good style, because they are more visible than single quotes.

There are four special data values: 

* `NA`: missing value, can be of any type

* `NaN`: "not a number", type **double** only; eg. `1 / 0` evaluates to `NaN`

* `Inf`: infinity, type **double** only

* `NULL`: the null object indicates absense of any object, and has no formal type

### Variable Assignment

To save for future use, **bind** them to names using either `<-` or `=`.

* `<-` is peferred "R style"; `->` for rightward assignment; <kbd>ALT</kbd> + <kbd>-</kbd> keyboard shortcut.

* `=` is preferred within function calls (more soon); leftward assignment only.

Name prohibitions:

* cannot begin with numbers or special characters
* beginning with `.` will create hidden objects

Style points:

* keep names as short and descriptive as possible.

* begin only with lowercase letters, eg. `treedata` not `Treedata`.

* use capital letters, `.`, and `_` to separate words or ideas.
    + eg. `treeData`, `tree.data`, `tree_data`. 
    
Two important points on variable names:

1. R is *always* case sensitive, eg. `treedata` and `treeData` are different!

2. whitespace *cannot* divide object names, numbers, or the assignment function.

  + eg. `tree data` or `20 17` both throw an **error** (more on errors soon).
  + `x <- 3` and  `x < - 3` give very different results!

### Functions

**Functions** are objects that invoke procedures to create new or modify existing objects

Functions have three important components:

1. a name used to specify the function to be called, and 

2. **formal arguments** to specify target objects and/or desired parameters of the procedure.

3. a **body** of code, accessable but hidden, usually written in R, but often in C.

The formal arguments of a function are named by the programmer, and must be specified if they do not have default values (but they often do!).

Access the *help file* to see a description of the function and its arguments: `?log`

To call a function, use this syntax: `function(arg1, arg2, arg3)`.

**Calling arguments** supplied to the function by the user, are matched to formal arguments first by name, then by position

* these are all the same call, in order from most to least explicit: `log(x=100, base=10)`, `log(base=10, x=100)`, `log(x=100, 10)`, `log(100, base=10)`, `log(base=10, 100)`, `log(100, base=10)`, `log(100, 10)`

* names will also match partially: `log(100, b=10)`, which is also good coding style, but I prefer `log(100, base=10)` 

* *always* use `=` to specify arguments, else the argument's object will be defined everywhere, not just within the function **environment**

* functions will use **default values** if they exist and the user does not specify their own values; see the help file

* often, you want to save the object **returned** by a function: `log.x <- log(100, base=10)`

* whitespace, even line returns, *can* separate components of a function call, which can be good style: 

```{r, eval=FALSE}

log (x = 100, 
     base = 10)

```
### Operators

Operators, like `+`, are **infix** functions (as opposed to prefix functions like `log()`), where the function call is *in*serted between two arguments.

Operators can be evaluated as objects by wrapping in ticks

#### Arithmetic Operators

* `+`, `-`, `*`, and `/` work as expected

*`^` is the power function

* use `(` function (and close with `)`) to specify non-standard order of operations

#### Logical Operators

The logical operators compare object values, and return logical vectors

Comparing numeric or character data:

* `>`, `<`, `>=`, `<=`, and `==` with *two* equal signs, unless

* prefix with `!` to negate; eg. `x != 2.5`

Comparing logical data:

* `&` for AND, `|` for OR, both elementwise
    + compares each element of left object to each element of right object
    + longer object must be multiple of shorter object; eg. `c(F, T, F) & T`
    + 0 coerced to FALSE, all other numeric data to TRUE
    + eg. use to determine if object is in range: `(x >= 0) & (x <= 1)`

* `&&` for AND, `||` for OR, compare left to right and return logical vector of length 1

### Errors and Warnings

The function programmers have built in requirements for function calls, and will return **error messages** when those requirements are not met.

* Errors indicate the function has failed to run, and will not return objects.

* Error messages vary in interpretabiity -- try to read and understand them, then start Googling

    + `log(base=10)`, `sqrt("a")`, and `c(T, F, T) | c(T, F)` give unusually helpful messages
    
Functions will also return **warning messages** if the programmer thinks the function output might not be as the user expects.

* Warnings indicate the function has run, and an object has been returned, but don't ingnore the warning! Sometimes could be very wrong, or not.

### Data Structures

#### Atomic Vectors

**Atomic vectors**: one-dimensional, homogenous value set of any one data type

Atomic vectors are the core data structure in R, so prioritize learning to create and maipulate them!

Creating atomic vectors: 

* `c()` combines the objects provided as calling arguments
    + can provide raw objects or variable names: eg. `c(1, 2, 3, x, y, z)`
* `seq()` creates a numerical sequence from and/or to user specified values with a given intervals and total length
* `:` operator creates integer sequence, eg. `0:10`, you will use this often!
* `rep()` repeats the elements of a vector; very flexible -- see formal arguments
* `rev()` reverses the elements of a vector

**Attributes** of atomic vectors:

* the elements of an atomic vector can be given **names** as a special attribute 
    + assign them inside the call to `c()`; eg. `c("a" = 1, "b" = 2, "c" = 3)`
    + or with `names()`; eg. `names(x) <- c("a", "b", "c")`.

**Coercion**: The elements of atomic vectors *must* be of the same type, and will be **coerced** to the most flexible type if not provided as such.

* flexibility: character > double > integer > logical

Assessing atomic vectors:

* `length()` will return the number of elements 
* `typeof()` will return the data type
* check on specific data types with `is.character()`, `is.numeric()`, `is.integer()`, `is.double()`, `is.logical()`
* `is.atomic` will check if an object is an atomic vector
* `names()` will return the character vector of element names

#### Factors

**Factors** are atomic vectors that can only take on pre-specified values, making them useful for representing categorical data.

Factors present to the user as character strings without quotes. But they are actually integer vectors, for which each integer value is associated with a character-string `level`, and the level printed instead of the integer. 

Creating factors:

* coerce atomic vectors with `as.factor()`, or use `factor()`
* factors are often created from character vectors by default in R -- this can be useful, but is a classic source of frustration!

Attributes of factors:

* factors have a `levels` attribute that is a character string of pre-determined values
* set or change the level values with `levels()`; eg. `levels(x) <- c("A", "B")`
* factors have a `class` attribute set to "factor" 

Assessing factors: 

* `is.factor()` or `class()` to determine if an object is a factor
* `levels()` returns levels attribute as character string
* `nlevels()` returns number of levels 

#### Lists

**Lists** are one-dimensonal vectors that can be heterogenous in data type

The elements of a list can themselves differ in data type, length, and dimensionality, and can be of any data structure (even lists!). Nevertheless, lists *per se* are 1D data structures.

Creating lists:

* `list()` combines elements into a list; used just like `c()`
* `rep()` and `rev()` work as expected 
* calling `c()` on multiple lists will combine their elements into one list
* flatten a list to an atomic vector using `unlist()`, but expect coersion!
* use `list()` or `names()` to assign names to each element as with atomic vectors

Assessing lists:

* use `str()` ("structure") to see the number of elements in a list, and the length and data type of each of its elements; very useful!!
* use `is.list()` or `typeof()` to determine whether an object is a list
* get names as with atomic vectors

#### Arrays and Matrices

**Arrays** are n-dimensional, homogenous value sets.

**Matrices** are a special array with two dimensions.

Creating arrays and matrices:

* `array()` and `matrix()` 
    + both take an atomic vector as a `data` argument.
    + `array()` takes an atomic integer vector to specify dimensions.
    + `matrix()` takes `nrow` and `ncol` arguments. 
    + both fill elements (rows) of each column first, unless `byrow = TRUE`.
    + both will repeat or omit elements of the data if dimensions do not match perfectly.
* combine atomic vectors, matrices, or arrays with `rbind()`, `cbind()`, and `abind()` for row-wise, column-wise, and multi-dimensional combination, respectively.
* transpose a matrix with `t()`, and permute an array with `aperm()`

Attributes of matrices and arrays:

* all matrices and arrays have a `dim` (dimension) attribute indicating the length of each dimension
* names can be assigned to rows and columns using `rownames()` and `colnames()`, or to any dimension using `dimnames()`

Assessing matrices and arrays:

* check dimensions with `dim()`, `nrow()`, `ncol()`. 
* check types with `typeof()`, `class()`, `is.matrix()` and `is.array()`.

#### Data Frames

**Data Frames** are lists of atomic vectors with the same length

Data frames present to the user as a matrix, not as a list, and character vectors are presented without quotes.

Attributes of Data Frames: 

* `names` is character string of column names; set with `names()` or `colnames()`
* `row.names` is character string of row names; set with `row.names()`
* `class` is set to "data.frame"

Creating Data Frames: 

* `data.frame()` performs column-wise combination of vectors, matrices, or data frames
    + add column names in function call; eg. `data.frame(y=0:10, x=seq(0, 100, by=10))`
* `cbind()` also does column-wise combination, but
    + at least one combined object must be a data frame, or else it will default to a coerced matrix
* `rbind()` performs row-wise combination, but
    + at least one combined object must be a data frame
    + column names of data frames and matrices must not conflict 

### Attributes

Attributes are supplementary properties applicable to all objects that provide object metadata.

Creating attributes:

* use `attr()` or `structure()`; eg, `attr(student1, "R knowledge") <- "rapidly growing"`

Assessing attributes: 

* `attr()` for specific attributes, or `attributes()` to see them all

Attributes are removed from objects when they are modified, except for the three special attributes already encountered:

* name attributes
* dimension attributes
* class attributes
