---
title: "Week 6 Homework"
author: "Tim Farkas"
date: "10/11/2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(results = "hold")
```

**Instructions**: Below are problems similar to the exercises done in class relating to the all course content up to and including Week 6. For questions that ask you to produce code, please insert one or a couple code chunks (using Insert button in shortcuts bar above, or on Mac <kbd>CMD</kbd> + <kbd>OPT</kbd> + <kbd>I</kbd>, on PC <kbd>CTL</kbd> + <kbd>ALT</kbd> + <kbd>I</kbd>). 

Put all code you want to execute inside a chunk. When you Knit your document, the code in chunks will be evaluated, and both the code and output will be printed. Any "inline" code (eg `x <- 10`) will not be evaluated, hence `x` in this case will not be created, and cannot be used by any downstream code. 

Please Knit to Word and upload a DOC file to Blackboard.

You can see more information about R Markdown in a Rmd file on Blackboard. 

In this homework, some pieces of code will not Knit properly if not run in the same chunk. Specifically, when you create a plot in a chunk, any further code modifying that plot needs in be in the same chunk where the plot was initially called. Furthermore, during development and testing, you will have run the whole chunk at once for modifications to execute properly (line by line will fail). Finally, if you want to visualize a plot in the Plot tab, you will need to  run the code in the Console (highlight, then CTL + RTN).

This will make some chunks rather lengthy, so please begin commenting your code by inserting a `#` on lines above the code text or after the code text on the same line. See Example Problem below. Any other comments that you think will be useful will be appreciated. Furthermore, please title your code chunks in the curly braces in the chunk header to indicate which problems are being addressed in each chunk. See the example below.

**Example Problem**

1. Create an integer vector `p` from 1 to 10 incrementing by 1. 
```{r Problem 1}
p <- 1:10 
```
2. Create an integer vector `prev` that reverses the elements of `p`
```{r Problem 2}
prev <- rev(p)
```
3. Given a vector `x`, make a vector `x2` for which each element is the square of the correponding element in `x`. 
4. Make a scatterplot relating `x` to `x2`.
5. Title the plot "Example Plot". 
```{r Problems 3-5}
# Problem E.3: Make Vectors.
x <- 1:10; x2 <- x ^ 2 

# Problem E.4: Create scatter plot of x against x2
plot(x, x2)

# Problem E.5. Add title to scatter plot
title(main="Example Plot") # axis labels left as default
```


**Preliminary**: Clear all objects from your global environment.

**Problems**

plant_height_data.csv has 60 records, where each record represents one plant, with 10 plants sampled at each of 6 sites. Half of the sites were fertilized twice per week after the first height measurement. Each plant was measured three times, one measurement per 10 days.

5.2. Data Import

1. Import the plant height data and save to `pd`. 
2. Remove any rows that have missing data. 
3. Check the structure of the data, including its dimensions and column data types. 

```{r}
pd <- read.csv("~/Dropbox/UNM R Course/plant_height_data.csv")
complete.cases(pd) # none!
str(pd)
```

6.2. Indexing Review (Optional)

Remember, you can extract elements of a data structure by passing three distinct data types to square brackets.

* positions, eg `pd[c(1,3,5), ]`,
* names, eg     `pd[ , c("m1", "m2", "m3")]`, or
* logicals, eg  `pd[ , c(F, F, F, F, T, T, T)]`

01. Extract a data frame `pd.m` from `pd` containing only the measurement variables `m1`, `m2`, and `m3`.
02. Extract all rows from `pd.m` for which values of `m1` are greater than 5.  
03. Extract all elements from `m2` for which values of `m1` are greater than 5.
04. Extract all rows from `pd.m` for which plants were fertilized.  
05. Make a histogram of `m1` values for which `m1` is less than 5 and `m2` is  less than 6.

```{r}
pd.m <- pd[ , c("m1", "m2", "m3")] #1
pd.m[pd.m$m1 < 5, ] #2
pd.m[pd.m$m1 < 5, "m2"] #3
pd.m$m2[pd.m$m1 < 5] #3 alternate
pd.m[pd$tmnt == "fert", ] # 4
hist(pd.m[pd.m$m1 < 5 & pd.m$m2 < 6, "m1"], #5
     main="", xlab="Plant Height at t = 0") # remove title, change x-axis label
```

6.3. Advanced Indexing

sev_lter_plant_spp.csv contains a catalogue of all the plant species in the Sevilleta Wildlife Refuge LTER primary productivity data. sev_col_oct17.csv contains records of plants collected by me this autumn for measurements of plant functional traits like plant height, stem density, and specific leaf area.

01. Create data frames for these data sets in R, calling them `spp` (plant species) and `col` (plants collected).
02. Create a vector `n.sp` of length == 2 with named elements indicating the number of species represented in each of the two datasets. 
03. Determine whether any species in the Oct 2017 are not in the list of LTER species. If there are some records of species in the Oct 17 data that are not in the LTER species list, remove those records from the Oct 17 data. Double check to make sure all species in the Oct 17 data are represented in the LTER species list.
04. Create an integer vector `sp.pos` indicating at what position (index) in the LTER data are each species represented in the Oct 2017 data. Hint: `?levels()`.
05. Check that each species in `col` has a match in `lter`. 

Oh, no! Removing all records of a factor level doesn't remove the factor level itself! 

06. Add a line of code in #3 to remove unused levels: `col <- droplevels(col)`.
07. Extract a data frame named `col.sp` from `lter` with only those rows of species in the Oct 17 data.
08. Use `table()` to count up how many records of each species are in the Oct 17 data. 
09. Add a column named `count` to `col.sp` indicating the number of records in the Oct 17 data for each species.
10. Sort `col.sp` by the number of records for each species, then by genus, then by species to break ties.
Bonus. Same as 10, but sort by number of records decreasing, and genus/species increasing.
11. Export the new, sorted data frame to your computer as a CSV. 

```{r}
lter <- read.csv("~/Dropbox/UNM R Course/sev_lter_plant_spp.csv") #1
col <- read.csv("~/Dropbox/UNM R Course/sev_col_oct17.csv") #1

n.sp <- c(LTER_spp = nlevels(lter$spp), col_spp = nlevels(col$spp)) #2

all(col$spp %in% lter$spp) #3.1
col <- droplevels(col[col$spp %in% lter$spp, ]) #3.2 and 6
all(col$spp %in% lter$spp) #3.3

sp.pos <- match(levels(col$spp), lter$spp) #4

any(is.na(sp.pos)) #5; not anymore, because I used droplevels() in #3.2 above.

col.sp <- lter[sp.pos, ] #7

table(col$spp) #8

col.sp$count <- table(col$spp) #9 only works if species are sorted in the data set and table of counts
all(col.sp$spp == names(table(col$spp))) # they are. phew!

# if they were not, use match!:

tab.match <- match(names(table(col$spp)), col.sp$spp)
col.sp$count[tab.match] <- table(col$spp)

col.sp <- col.sp[order(col.sp$count, col.sp$genus, col.sp$species), ] # 10
write.csv(col.sp, "PATH/name.csv")

col.sp <- col.sp[order(col.sp$genus, col.sp$species), ] # Bonus Step 1
col.sp <- col.sp[order(col.sp$count, decreasing=TRUE), ] # Bonus Step 2

```

5.4. Data Manipulation

01. Import plant_height_data.csv, and save it as a data frame named `pd`.
02. Split `pd` into a list of data frames named `pd.l` each with records from only one site.
03. Convert the data for site B from wide to long format, stacking `m1`, `m2`, and `m3` on top of one another, preserving `id` and `tmnt` as identifier variables. Save as `pd.B.long`.
04. Remove all records for which id == 20. 
05. Convert `pd.B.long` back to wide format (without records for which id == 20), calling it `pd.B.wide`.
06. Overwrite the element in `pd.l` for site B with the new wide-format data frame.
Bonus. Unsplit `pd.l`. Good luck ... 

```{r}
pd <- read.csv("~/Dropbox/UNM R Course/plant_height_data.csv") #1
pd.l <- split(pd, f = pd$site) #2
library(reshape) #3
pd.B.long <- melt(pd.l$B, id.vars = c("id", "tmnt"), measure.vars = c("m1", "m2", "m3"))
pd.B.long <- pd.B.long[pd.B.long$id != 20, ]
pd.B.wide <- cast(pd.B.long, form = id + tmnt ~ variable)
pd.l$B <- pd.B.wide
```

5.5. Function Building (Optional)

01. Make a function `myhist()` that takes a vector of numeric data and prodcues a histogram with the number of bars equal to the length of the vector divided by a user-provided integer, with a default of 10. Test it. 

```{r}
myhist <- function(x, div = 10) {
  
  bks <- length(x) %/% div # integer division to round down to nearest integer
  hist(x, breaks = bks)
  
}

myhist(rnorm(100)) # hmmm, not quite right. see ?hist.
myhist(rnorm(100), div=2) # yeah, ok, it seems to be working ish

```


