---
title: "Week 4 Homework"
author: "Tim Farkas"
date: "9/12/2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
#knitr::opts_chunk$set(results = "hold")
```

**Instructions**: Below are problems similar to the exercises done in class relating to the all course content up to and including Week 4. For questions that ask you to produce code, please insert one or a couple code chunks (using Insert button in shortcuts bar above, or on Mac <kbd>CMD</kbd> + <kbd>OPT</kbd> + <kbd>I</kbd>, on PC <kbd>CTL</kbd> + <kbd>ALT</kbd> + <kbd>I</kbd>). 

Put all code you want to execute inside a chunk. When you Knit your document, the code in chunks will be evaluated, and both the code and output will be printed. Any "inline" code (eg `x <- 10`) will not be evaluated, hence `x` in this case will not be created, and cannot be used by any downstream code. 

Please Knit to Word and upload a DOC file to Blackboard.

You can see more information about R Markdown in a Rmd file on Blackboard. 

In this homework, some pieces of code will not Knit properly if not run in the same chunk. Specifically, when you create a plot in a chunk, any further code modifying that plot needs in be in the same chunk where the plot was initially called. Furthermore, during development and testing, you will have run the whole chunk at once for modifications to execute properly (line by line will fail). Finally, if you want to visualize a plot in the Plot tab, you will need to  run the code in the Console (highlight, then CTL + RTN).

This will make some chunks rather lengthy, so please begin commenting your code by inserting a `#` on lines above the code text or after the code text on the same line. See Example Problem below. Any other comments that you think will be useful will be appreciated. Furthermore, please title your code chunks in the curly braces in the chunk header to indicate which problems are being addressed in each chunk. See the example below.

**Example Problem**

1. Create an integer vector `p` from 1 to 10 incrementing by 1. 
```{r Problem 1}
p <- 1:10 
```
2. Create an integer vector `prev` that reverses the elements of `p`
```{r Problem 2}
prev <- rev(p)
```
3. Given a vector `x`, make a vector `x2` for which each element is the square of the correponding element in `x`. 
4. Make a scatterplot relating `x` to `x2`.
5. Title the plot "Example Plot". 
```{r Problems 3-5}
# Problem E.3: Make Vectors.
x <- 1:10; x2 <- x ^ 2 

# Problem E.4: Create scatter plot of x against x2
plot(x, x2)

# Problem E.5. Add title to scatter plot
title(main="Example Plot") # axis labels left as default
```


**Preliminary**: Clear all objects from your global environment.

**Problems**

4.1. Given `x <- rnorm(n=100, mean=5, sd=10)` produce a histogram of those values in `x`

1. in `x` (all of them!)
2. less than the mean of the sampled normal distribution. Hint: See call to `rnorm()` above, and look for the word "mean". Also, `?rnorm`.
3. less than the mean of the sample (ie `x`)
4. less than 3 and greater than 7
5. Do the histograms for 4.1.2 and 4.1.3 differ? Why or why not? 

Bonus 6. Make histograms look nicer, with appropriate axis labels, a good title, your preferred bin number, some color, and perhaps some curves to show the distributions of data.

4.2. Given the vector of data `y <- rpois(100, 100)`

01. create a matrix with 10 rows, filling column by column.
02. create a matrix `ymat` with 20 rows, filling by row.
03. name the rows of `ymat` 1 to 20, and columns A through E (see `?LETTERS`)
04. predict `t(t(ymat))`.
05. extract the element in the 5th row, 2nd column.
06. extract the element in the 2nd row, 5th column. 
07. extract all elements in the 2nd row. 
08. extract all elements in the 5th column, returning a matrix.
09. extract all elements in the 2nd and 5th column.
10. extract the elements in all rows, except rows 2 and 5.
11. extract the elements in rows 5 to 15, inclusive.
12. extract the elements in all rows, excluding rows 5 to 15.


4.3. Given `ymat` from problem 4.2, create a numeric vector of all elements in `ymat`:

1. less than the mean of its elements
2. less than 90 or greater than 110
3. less than 90, greater than 110, or equal to 100
4. divisible by 2
5. less than 90 or greater than 110, and also divisible by 2

4.4. Given `ymat` from problem 4.2, extract
 
1. the rows (all columns) for which the element in column 1 is greater than 100
2. the columns (all rows) for which the element in row 5 is either less than 95 or greater than 105. Be sure to return a matrix.
 
The functions `colSums()`, `rowSums()`, `colMeans()`, and `rowMeans()` efficiently compute the sums or means of all rows or columns in matrices and data frames.

3. extract the rows for which the sum of their elements is greater than 500.
4. extract the columns for which the mean of their elements is less than 100.

4.5. Using the `iris` data frame, 

1. create a new data frame `iris.se` containing only records for *Iris setosa*, and removing the column indicating the species
2. make a histogram of sepal length for *Iris setosa*
3. make a named vector giving the means for each column in `iris.se`
4. extract those rows from `iris.se` for which sepal length is less than the mean (or sepal length)
5. extract petal width measurements from `iris.se` for which sepal length is less than the mean (of sepal length)
6. extract all rows from `iris.se` for which sepal length is less than the mean (of sepal length) and petal length is greater than the mean (of petal length)





